---
title: "Testing Additivity of Lead and Benzo[a]pyrene--induced Neurotoxicity in *Caenorhabditis elegans* Assays"
author: "Niccolò Anceschi"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", cache = FALSE
)
```

# Overview

This document reproduces the main results for the BaP–Pb toxicant additivity analysis 
from the manuscript [Anceschi, Huayta, Meyer, Dunson and Herring, (2026+)
"*Testing Additivity of Lead and Benzo[a]pyrene--induced Neurotoxicity in \textit{Caenorhabditis elegans} Assays*"](https://arxiv.org/pdf/PLACEHOLDER).
It loads the pre-processed *C. elegans* assay data, fits the single-exposure profiles and
the joint radial-spline surface, and producess figures for the dose-response surfaces
and log-likelihood comparisons reported in the paper. 
Each section below sources one script from the pipeline, in dependency order.
Each sourced script contains dedicated and detailed documentation of its contents.

The convex-mixture model employed represents every ordinal damage-score distribution as
$(1-\lambda)F_0 + \lambda F_\infty$, interpolating between an unexposed and a maximal-effect profile. 
The dose dependence enters entirely through the weight $\lambda(\cdot)$, 
a monotone non-decreasing function that governs the whole outcome distribution.
Under joint exposure, it becomes a surface $\lambda_{12}(d_1,d_2)$ over the dose plane.
Additivity is tested by a likelihood ratio comparing the fitted joint model
against different nulls -- Bliss, Loewe$^+$, and Hand -- 
each inducing an additive baseline surface $\lambda_{12}^{(o)}$ built from the single-exposure fits.

# Configuration

When this document is opened in RStudio, the repository path is detected automatically from the file's location, so no manual editing is needed.
`data_file` names the pre-processed data expected alongside the scripts.

```{r config}
source_dir <- dirname(rstudioapi::getSourceEditorContext()$path) # repo local path 
data_file <- "C-elgans_preprocessed_data.rds" # pre-processed data file
```

# Utilities

Load required packages and define generic helpers (softmax, mixture
probabilities, monotone inversion by root-finding).

```{r utils, message=FALSE, warning=FALSE}
source(file.path(source_dir, "Utils.R"))
```

# Data import and setup

Import the pre-processed assay data and build everything that does *not* depend on
the fitted parameters: the single- and joint-exposure spline bases, the dose
constants and spline dimensions, the extremal-profile PMFs, and nested
exposure data subsets.

```{r data}
source(file.path(source_dir, "Data_setup.R"))
```

# Likelihoods and radial splines

Define the likelihood functions — both for single and joint exposures — and the
penalized radial-spline objective — together with the caching helpers that keep
each optimizer step cheap.

```{r loglik}
source(file.path(source_dir, "Loglikelihood.R"))
```

# Additive null models

Define the three additive baselines $\lambda_{12}^{(o)}(d_1,d_2)$ — Bliss, Loewe$^+$
(EME), and Hand — using spline parameters to be fitted on single-exposure data.

```{r null-models}
source(file.path(source_dir, "Null_models.R"))
```

# Model fit

Fit the two stages: first the single-exposure behavior for both BaP and Pb,
then the joint-exposure surface via radial-splines under a monotonicity penalty. 
This is the computational core; the timing messages from `tic()`/`toc()` report each fit.

```{r fit, cache=TRUE}
source(file.path(source_dir, "Model_fit.R"))
```

# Result Visualization

Produce the fitted weight-function plots, the joint weight-surface contours for
each null and for the radial fit, the fitted-vs-empirical CDF scatterplots, and
the across-subset log-likelihood comparison.

```{r plots-source, message=FALSE}
source(file.path(source_dir, "Plots.R"))
```

Single-exposure weight functions and the Pb→BaP dose alignment.

```{r fig-weights, fig.width=9, fig.height=3, out.width="100%", dpi=250}
plot_weight_functions(save = FALSE)
```

Joint weight surfaces: the three additive nulls and the radial fit.

```{r fig-surfaces, fig.width=8.5, fig.height=8, out.width="90%", dpi=150, cache=TRUE}
plot_weight_surfaces(save = FALSE)
```

Log-likelihood across data subsets and models.

```{r fig-loglik, fig.width=7.5, fig.height=3.5, out.width="70%", dpi=250}
plot_loglik(save = FALSE)
```

Fitted vs. empirical CDFs, all data and joint exposures only.

```{r fig-cdf, fig.width=9, fig.height=5, out.width="80%", dpi=250}
plot_cdf_scatter(save = FALSE)
```

# Bootstrap Test for Additivity

The additivity tests are calibrated by a parametric conditional bootstrap:
for each null model, replicate datasets of the joint exposures data are drawn
from the corresponding additive baseline, the radial surface is refitted on 
each, and the likelihood-ratio statistic is recomputed to build its null
distribution -- from which p-values and power are obtained.

Refitting the model on every replicate is computationally intensive:
the analysis in the manuscript used 10,000 replicates per hypothesis and was run
in parallel across multiple nodes on a computing cluster.

For transparency and completeness, the full procedure -- replicate simulation, 
aggregation, and computation of the test statistics, p-values, and power -- is provided in `Bootstrap_test.R`.




